% ======================================================================
% 00_info-gas.tex — Relational info–gas on boundaries
% ======================================================================

\subsection{Info–Gas \& Stokes}
\label{corollary:info-gas}

\paragraph{Literature scope.}
All corollaries below follow from the generalized Stokes identity \(d(*j)=\sigma\,\mathrm{vol}\) and the
\emph{frame} linear closure \(\mathbf j=-\kappa\,\mathbf E_I\), with optional Info--EM specialization where stated.
Standard identities (e.g., Poynting) are recalled in Apps.~\ref{app:forms-stokes}--\ref{app:jacobson} and used only
when explicitly invoked in the relevant corollary.

\paragraph{Entropic Flux.}
Boundary counting at an instant is represented by the \emph{entropic flux} \(1\)-form \(j\) on \(\partial U\).
On a \emph{frame}, its normal component \(j_n:=\mathbf j\!\cdot n\) is the boundary flux density
(units: count\,/\,area\,/\,time; the log base fixes the count unit, e.g.\ nats or trits).
The instantaneous flux through a boundary patch \(A\subset\partial U\) is
\[
\Phi_A(t):=\int_A j_n(\cdot,t)\,dA .
\]
All tube/boundary invariants in this section are \emph{frame} statements.

\paragraph{Entropic Current.}
Over a time window \([t,t+\Delta t]\) on a boundary patch \(A\), define the time–integrated boundary
\emph{counts}
\[
N_A(\Delta t):=\int_t^{t+\Delta t}\!\!\int_A j_n\,dA\,dt,
\]
the window–average \emph{current} (rate) \(I_A:=N_A(\Delta t)/\Delta t\), and the window–average \emph{current density}
\(\bar j_n:=\frac{1}{\Delta t}\!\int_t^{t+\Delta t} j_n\,dt\).
Window fluctuations are summarized by the Fano factor \(F_A:=\mathrm{Var}[N_A]/\mathbb E[N_A]\) and an effective
integration time \(\Delta t_{\!\mathrm{eff}}\le\Delta t\) (from the count autocorrelation). These window notions power
the time–information bounds and estimator floor (C5/C6).

\paragraph{Boundary bookkeeping (frame vs.\ window).}
\emph{Assumption (open set).} Work on \(W\subset U\) with \(\sigma|_W\equiv 0\) (source–free on \(W\)).
On a \emph{fixed region} \(U\) (a frame), the cut identity is
\begin{equation}
\dot I_{\mathrm{enc}}=\Phi_{\mathrm{in}}-\Phi_{\mathrm{out}}+\Pi,
\qquad
\Phi_{\mathrm{in/out}}=\int_{\partial U}(\mathbf j\!\cdot n)_{\mp}\,dA,
\qquad
\Pi=\int_{U}\sigma\,d^n x ,
\label{corollary:info-gas:eq:cut}
\end{equation}
which is \eqref{axioms:stokes:identity:eq:cut} in the axioms. The \emph{frame} linear closure is
\begin{equation}
\mathbf E_I:=-\nabla\Phi_I,
\qquad
\mathbf j=-\kappa\,\mathbf E_I ,
\label{corollary:info-gas:eq:frame-closure}
\end{equation}
identical to \eqref{axioms:stokes:frame:eq:closure}.

\paragraph{Tube invariance (frame, source–free open set).}
Consider a narrow tube \(\mathcal T\) whose side is everywhere tangent to \(\mathbf j\).
For any cross–section \(\Sigma(r)\) orthogonal to the tube axis, define the directed flux
\begin{equation}
\Phi(r):=\int_{\Sigma(r)}\mathbf j\!\cdot n\,dA .
\label{eq:c0-flux}
\end{equation}
If \(\partial_t i=0\) on the frame and if \(\mathcal T\subset W\subset U\) with \(\sigma|_W\equiv 0\),
then \(d j^{(n)}=0\) implies constant tube flux,
\begin{equation}
\Phi(r)=\mathrm{const},\qquad
\big\langle j_n\big\rangle(r)=\frac{\Phi}{A(r)},
\label{corollary:info-gas:eq:tube}
\end{equation}
where \(A(r):=\int_{\Sigma(r)}dA\) and \(\langle j_n\rangle\) is the cross–sectional mean of the normal component.

\paragraph{Cones \& boundary sectors (dimension–dependent scaling).}
For a conical boundary sector of fixed solid angle \(\Delta\Omega\) in \(n\) spatial dimensions,
\begin{equation}
A_n(r)=\Delta\Omega\,r^{\,n-1}\quad\Rightarrow\quad
\big\langle j_n\big\rangle(r)=\frac{\Phi}{\Delta\Omega}\,\frac{1}{r^{\,n-1}} .
\label{corollary:info-gas:eq:cone}
\end{equation}
With \eqref{corollary:info-gas:eq:frame-closure},
\begin{equation}
\big|\mathbf E_I(r)\big|
=\frac{1}{\kappa}\,\big\langle j_n\big\rangle(r)
=\frac{1}{\kappa}\,\frac{\Phi}{A_n(r)}
\ \propto\ r^{-(n-1)} .
\label{corollary:info-gas:eq:sphere}
\end{equation}
\[
n=1:\ \big|\mathbf E_I\big|\propto r^{0}\ (\text{no geometric decay}),\qquad
n=2:\ \big|\mathbf E_I\big|\propto r^{-1},\qquad
n=3:\ \big|\mathbf E_I\big|\propto r^{-2}.
\]
The \(n{=}1\) case captures confinement in tubes/guides (constant cross–section), while \(n{=}2\) captures
sheet/cylindrical spreading. These are revisited and generalized with waves in C8 (\S\ref{corollary:dimensional});
for wave export/envelopes, see §\ref{waves} (cf.\ eq.~\eqref{waves:eq:envelope}).

\paragraph{Directional filtering.}
For apertures or one–sided diaphragms, replace the total flux by the inward component
\(\Phi_-(r)=\int_{\Sigma(r)}(\mathbf j\!\cdot n)_-\,dA\).
Formulas \eqref{corollary:info-gas:eq:tube}–\eqref{corollary:info-gas:eq:sphere} hold with \(\Phi\mapsto\Phi_-\).

\paragraph{Heat (sector mapping).}
Temperature is not part of the geometric law. If energy accounting is desired, map entropic quantities to heat:
on a frame, the heat–flux density is \(q_n(t)=T(t)\,j_n(t)\); over a window,
\[
Q_A=\int_t^{t+\Delta t}\!\!\int_A T\,j_n\,dA\,dt
\quad(\text{if }T\text{ is uniform on the window/patch, }Q_A\approx T\,N_A).
\]
If \(T\) varies, one may use the entropy–weighted temperature
\(T^\ast := \frac{\int\!\!\int T\,j_n}{\int\!\!\int j_n}\) so \(Q_A=T^\ast N_A\).
This Clausius mapping is recorded in \S\ref{corollary:continuity} \S\ref{corollary:info-gas} remains temperature–free.

\paragraph{Trits, nats, and units.}
Recording counts in \emph{trits} maps to nats via \(j_{\rm nats}=(\ln 3)\,j_{\rm trits}\).
Equations \eqref{corollary:info-gas:eq:tube}–\eqref{corollary:info-gas:eq:sphere} are invariant under log–base choice;
only the numerical value of \(\kappa\) rescales accordingly. When a measured sector is desired, a single datum in
Kz (SI$/h$) fixes the linear calibration (App.~\ref{app:calibrations}).

\paragraph{What this establishes.}
C0 isolates the \emph{surface (area), not line} nature of conservation on boundaries:
\emph{flux} \(\Phi\) is invariant along tubes, while \emph{amplitudes} scale as \(1/A\).
Use \emph{Entropic Flux} for frame relations (inverse–area laws, tube invariants) and \emph{Entropic Current}
for window relations (counts, rates, and their statistics). Heat appears only via the optional boundary Clausius
map when energy units are required.
