% ======================================================================
% 03c08_dimensional-flows-confinement.tex — C8: Dimensional flows & confinement
% ======================================================================

\subsection{C8: Dimensional flows \& confinement}
\label{corollary:dimensional}
This corollary makes explicit how static and wave amplitudes scale with the number of accessible spatial
dimensions and how boundaries reduce the \emph{effective} dimension. All statements follow from
the Stokes identity \eqref{eq:stokes-identity} (D1), the static closure \eqref{eq:static-closure} (D2), and
(for waves) the exported–power identity \eqref{eq:poynting} (App.~\ref{app:forms-stokes}; see also C4,
\S\ref{corollary:info-em}). No new constants are introduced.

% ------------------------- C8.0 — Summary (static vs waves) ------------

\paragraph{C8.0 \; Summary (static vs.\ waves).}
Let \(A_n(r)\) be the area of a radius–\(r\) screen in \(n\) spatial dimensions (App.~\ref{app:nD-screens},
eq.~\eqref{appC:eq:areas}). On a steady, source‑free annulus the tube/screen flux \(\Phi\) is constant (C0:
\S\ref{corollary:info-gas}), so the static mean and amplitude follow (C1: \S\ref{corollary:static-screen})
\begin{equation}
\big\langle j_n\big\rangle(r)=\frac{\Phi}{A_n(r)},\qquad
\big|\mathbf E_I(r)\big|=\frac{1}{\kappa}\,\frac{\Phi}{A_n(r)} .
\label{eq:c8-static}
\end{equation}
For spherical/conical screens with \(A_n(r)=S_{n-1}\,r^{\,n-1}\) (App.~\ref{app:nD-screens}, eq.~\eqref{appC:eq:areas}),
this gives \(|\mathbf E_I|\propto r^{-(n-1)}\).
For wave segments in homogeneous media, time‑averaging the Poynting identity on a source‑free annulus yields
radius‑independent exported power (C4: \S\ref{corollary:info-em}),
\begin{equation}
P(r):=\int_{\Sigma(r)}\!\big\langle \mathbf S_I\big\rangle\!\cdot n\,dA=\text{const in }r ,
\label{eq:c8-power}
\end{equation}
and with \(\langle \mathbf S_I\rangle\propto \langle|\mathbf E_I|^2\rangle\) the far‑field envelope scales as
\begin{equation}
\big\langle |\mathbf E_I(r)| \big\rangle \;\propto\; \frac{1}{\sqrt{A_n(r)}}
\quad\text{(cf.\ C4, eq.~\eqref{eq:c4-amp-scaling})}.
\label{eq:c8-waves}
\end{equation}
For spherical/conical screens this gives \(\langle|\mathbf E_I|\rangle\propto r^{-(n-1)/2}\).

% ------------------------- C8.1 — Dimensional table --------------------

\paragraph{C8.1 \; Dimensional summary (table).}
\begin{table}[t]
\centering
\caption{Geometric scaling of static amplitudes and wave envelopes by effective spatial dimension.
Here $A_n(r)$ is the $n$D screen area (App.~\ref{app:nD-screens}). Static scaling follows \eqref{eq:c8-static};
wave envelopes follow \eqref{eq:c8-power}–\eqref{eq:c8-waves}.}
\label{tab:c8-dim}
\begin{tabular}{lccc}
\hline
\textbf{Effective dimension} & \textbf{Area $A_n(r)$} & \textbf{Static $|\mathbf E_I|$} & \textbf{Wave $\langle|\mathbf E_I|\rangle$} \\
\hline
1D (guide / tube)     & $A_\perp$ (const.)     & $\propto r^{0}$   & $\propto r^{0}$ \\
2D (sheet / interface)& $A_2(r)\propto r$      & $\propto r^{-1}$  & $\propto r^{-1/2}$ \\
3D (spherical)        & $A_3(r)=4\pi r^{2}$    & $\propto r^{-2}$  & $\propto r^{-1}$ \\
\hline
\end{tabular}
\end{table}

\noindent\emph{Sector note.} For a fixed solid-angle sector $\Delta\Omega$ in $n$D,
replace $A_n(r)$ by $A_\Delta(r)=\Delta\Omega\,r^{\,n-1}$ (App.~\ref{app:nD-screens}); the exponents are unchanged.

% ------------------------- C8.2 — Confinement cases --------------------

\paragraph{C8.2 \; Confinement cases (effective dimension).}
Boundaries that constrain propagation reduce the accessible dimension \(n_{\rm eff}\):
\begin{itemize}\itemsep2pt
\item \textbf{Waveguides (effective 1D).} Cross–section \(A_\perp\) fixed along the axis.
Static: \(\langle j_n\rangle=\Phi/A_\perp\) and \( |\mathbf E_I|=\Phi/(\kappa A_\perp)\) are independent of \(r\).
Waves: \(\langle|\mathbf E_I|\rangle\sim r^0\) along the guide; mode power is exported with \eqref{eq:c8-power}.
\item \textbf{Sheets / interfaces (effective 2D).} Fixed thickness \(\Rightarrow A_2(r)\propto r\).
Static: \( |\mathbf E_I|\propto 1/r\). Waves: \( \langle|\mathbf E_I|\rangle\propto 1/\sqrt{r}\).
\end{itemize}
Mode spectra and group velocities are set by boundary conditions and any chosen constitutive map \(\chi\).

% ------------------------- C8.3 — Directional sectors ------------------

\paragraph{C8.3 \; Directional sectors and apertures.}
For an aperture subtending \(\Delta\Omega\) (in \(n\)D), use \(A_\Delta(r)=\Delta\Omega\,r^{\,n-1}\) in
\eqref{eq:c8-static}–\eqref{eq:c8-waves}. This is the sector (tube) form of the same conservation rule
(C0: \S\ref{corollary:info-gas}).

% ------------------------- C8.4 — Remarks & scope ----------------------

\paragraph{C8.4 \; Remarks.}
(i) Near–field structure depends on sources and boundary detail; the listed wave scalings are far–field envelopes.\\
(ii) Edge diffraction in 2D sheets and leakage in imperfect guides enter through attenuation/survival factors (C11:
\S\ref{corollary:entropic-corrections}) and modify only prefactors, not the exponents, in the asymptotic limits.

\medskip
\noindent\emph{Summary.}
Static amplitudes fall as \(1/A_n(r)\); wave envelopes fall as \(1/\sqrt{A_n(r)}\). Confinement replaces \(A_n(r)\)
by the appropriate effective cross–section or circumference. These follow directly from the Stokes identity,
the static closure, and constant exported power, with no additional postulates.

\medskip
\noindent\emph{Literature note.}
Background geometry/areas and cone/tube forms are summarized in App.~\ref{app:nD-screens};
the Poynting/export statement is recalled in App.~\ref{app:forms-stokes} and C4. For standard expositions,
see Frankel~\cite{Frankel2011} (geometry) and Jackson~\cite{Jackson1999} or
Landau--Lifshitz--Pitaevskii~\cite{LandauLifshitzEDCM1984}.
