% ======================================================================
% 03c09_thermo-newton-laws.tex — C9: Thermodynamics → Newtonian rate laws
% ======================================================================

\subsection{C9: Thermodynamics to Newtonian rate laws (constant‑free)}
\label{corollary:thermo-newton}

We show that the Newtonian update $\dot{\mathbf p}=\mathbf f$ and the work–power identity follow from:
(i) the canonical rate equations obtained by evaluating Stokes on a rectangular $2$‑surface in $(q,t)$,
(ii) the static informational closure for the surrounding channel, and
(iii) a single, sector‑specific calibration (App.~\ref{app:calibrations}). No bare constants are introduced.

\paragraph{Setup: canonical rates from Stokes.}
From the rate $1$‑form (Axiom D4),
\begin{equation}
\Theta_{\mathcal R}(q,P,t)=P\!\cdot dq - H(q,P,t)\,dt,
\qquad
d\Theta_{\mathcal R}=0 ,
\label{eq:c9-theta}
\end{equation}
Stokes on a rectangular $2$‑surface spanned by two nearby paths yields the canonical rate equations
\begin{equation}
\dot q = \partial_{P}H,
\qquad
\dot P = -\,\partial_{q}H ,
\label{eq:c9-canonical}
\end{equation}
matching \eqref{eq:pc-form}–\eqref{eq:canonical}.

\paragraph{Minimal kinetic choice and Newton’s update.}
Choose a quadratic kinetic term and an external driving potential,
\begin{equation}
H(q,P,t)=\tfrac12\,P\!\cdot M^{-1}P \;+\; U(q,t),
\label{eq:c9-kinetic}
\end{equation}
with $M$ a constant (or slowly varying) positive‑definite inertia matrix.
Then \eqref{eq:c9-canonical} gives
\begin{equation}
\dot q = M^{-1}P \quad\Rightarrow\quad P=M\,\dot q \;(:=\mathbf p),
\qquad
\dot P = -\,\partial_q U \;(:=\mathbf f),
\label{eq:c9-newton}
\end{equation}
i.e.\ the Newtonian rate law $\dot{\mathbf p}=\mathbf f$.
If $M=M(q)$, standard geometric terms appear in $\dot{\mathbf p}$ via $-\partial_q\!\bigl(\tfrac12 P\!\cdot M^{-1}P\bigr)$.

\paragraph{Work–power identity.}
Let $K(P):=\tfrac12 P\!\cdot M^{-1}P=\tfrac12\,\dot q\!\cdot M\dot q$.
Then, using \eqref{eq:c9-newton},
\begin{equation}
\frac{dK}{dt}
= \dot q\!\cdot \dot P
= \dot q\!\cdot \mathbf f
= \mathbf f\!\cdot \mathbf v ,
\qquad \mathbf v:=\dot q ,
\label{eq:c9-power}
\end{equation}
so the instantaneous power delivered by the driving equals the kinetic‑energy rate.
If $U$ has explicit time dependence, $\dot H=\partial_t U$ with $H:=K+U$; for time‑independent $U$,
$\dot H=0$ and $\,\mathbf f\!\cdot\mathbf v=-\,\dot U$.

\paragraph{Entropic/operational reading (via REOS).}
In the REOS language (Axiom D3), boundary/control variations $R^{(a)}$ with price $\lambda_a$ satisfy
$d\rho=-\lambda_a\,dR^{(a)}$. When a control field induces a spatial price gradient,
the force on a probe reads as the $q$‑gradient of a potential, i.e.\ $\mathbf f=-\partial_q U$ in
\eqref{eq:c9-newton}. Non‑integrable control cycles ($d\lambda\neq0$) incur nonnegative production
$\sigma\ge0$, which registers via \eqref{eq:stokes-identity}–\eqref{eq:cut-balance}.

\paragraph{Notes and variants.}
\begin{itemize}\itemsep2pt
\item \emph{Vector/tensor inertia.} For anisotropic agents, $M$ is a matrix; \eqref{eq:c9-newton} gives
$\mathbf p:=M\,\dot q$ and $\dot{\mathbf p}=\mathbf f$ unchanged (with the usual Christoffel‑type terms if $M(q)$).
\item \emph{Other sectors.} The same construction carries to other sectors: replace $U(q,t)$ with the sectoral
potential appropriate to the chosen readout (gravitational, electric, thermal) and read off $\mathbf f=-\partial_q U$.
\item \emph{Dissipation.} Adding a Rayleigh term $\mathcal R(\dot q)$ to the rate generator replaces
$\dot P=-\partial_q U$ by $\dot P=-\partial_q U-\partial_{\dot q}\mathcal R$, producing a drag consistent with
nonnegative production $\sigma$ in \eqref{eq:stokes-identity}–\eqref{eq:cut-balance}.
\end{itemize}

\medskip
\noindent\emph{Summary.}
A minimal kinetic choice in the Stokes‑evaluated canonical system yields $\dot{\mathbf p}=\mathbf f$ and the
work–power identity in a constant‑free form. Sectoral forces arise as $-\partial_q U$; dissipative additions
match the production term captured by \eqref{eq:stokes-identity}–\eqref{eq:cut-balance}. The same pattern
transfers to other static sectors by changing only the sectoral potential $U$.

\medskip
\noindent\emph{Literature note.}
The Poincar\'e–Cartan form and Hamiltonian rates are standard (see Arnold~\cite{Arnold1989}); convex duality
and Legendre regularity are summarized in Rockafellar~\cite{Rockafellar1970}. Our derivation uses only the
Stokes evaluation on $(q,t)$ (Axiom D4) and one‑datum sector calibration (App.~\ref{app:calibrations}).
