% ======================================================================
% 03c06_noise-floor.tex — C6: Noise floor ("force cloud")
% ======================================================================

\subsection{C6: Noise floor (``force cloud'')}
\label{corollary:noise}

This corollary quantifies the estimator floor on the informational field from boundary counting
and shows how it propagates to calibrated sectors. Let $A$ be a readout patch with outward unit
normal $n$, observed over a window $[t,t{+}\Delta t]$. Define the inward flux and the patch‑mean
normal current as
\begin{equation}
J_A \;:=\; \int_A \big(\mathbf j\!\cdot n\big)_{-}\, dA,
\qquad
\overline{j}_A \;:=\; \frac{J_A}{|A|}.
\label{eq:c6-flux-defs}
\end{equation}

\paragraph{Windowed counting witness (definitions and operational bound).}
Let $N_A(\Delta t)$ be the inward counts on $A$ over the window $\Delta t$. Write the mean count as
\begin{equation}
\mu \;:=\; \mathbb E[N_A]\;=\; J_A\,\Delta t ,
\label{eq:c6-poisson-mean}
\end{equation}
and the \emph{windowed Fano factor} as
\begin{equation}
\mathrm{Var}\big(N_A\big)\;=\;F_A\,\mu .
\label{eq:c6-crb-J}
\end{equation}
% ---- Neutral alias for backward compatibility (same equation number)
\begingroup
\makeatletter
\edef\@currentlabel{\theequation}%
\makeatother
\label{eq:c6-witness-mean}
\endgroup
% ----------------------------------------------------------------------

If arrivals exhibit temporal correlation with effective correlation time $\tau>0$, define an
\emph{effective window} $\Delta t_{\!\mathrm{eff}}\le \Delta t$ from the count autocorrelation on
the same record (App.~\ref{app:noise-kernel}). Any boundary estimator of the patch‑mean current
admits the operational benchmark
\begin{equation}
\mathrm{Std}\!\big(\widehat{\overline{j}}_A\big)
\;\approx\;
\sqrt{\frac{F_A\,\overline{j}_A}{|A|\,\Delta t_{\!\mathrm{eff}}}} .
\label{eq:c6-crb-jbar}
\end{equation}
% ---- Neutral alias for backward compatibility (same equation number)
\begingroup
\makeatletter
\edef\@currentlabel{\theequation}%
\makeatother
\label{eq:c6-witness-jbar}
\endgroup
% ----------------------------------------------------------------------

Shot‑noise–limited records have $F_A\!\approx\!1$ and $\Delta t_{\!\mathrm{eff}}\!\approx\!\Delta t$;
overdispersed or correlated arrivals are captured by the measured $(F_A,\Delta t_{\!\mathrm{eff}})$.

\paragraph{From $\mathbf j$ to the informational field $\mathbf E_I$ (snapshot mapping).}
Using the static closure \eqref{eq:static-closure}, $\ \mathbf j=-\kappa\,\mathbf E_I$, the \emph{windowed}
(local) floor for $\mathbf E_I$ is
\begin{equation}
\boxed{\quad
\mathrm{Var}(\widehat{E}_I)\ \gtrsim\ \frac{F_A\,\overline{j}_A}{\kappa^{2}\,|A|\,\Delta t_{\!\mathrm{eff}}},
\qquad
\sigma_{E_I}(\Delta t)\ \gtrsim\ \frac{1}{\kappa}\,
\sqrt{\frac{F_A\,\overline{j}_A}{|A|\,\Delta t_{\!\mathrm{eff}}}} \ .
\quad}
\label{eq:c6-var-EI}
\end{equation}
In the shot‑noise limit ($F_A\!\approx\!1$, $\Delta t_{\!\mathrm{eff}}\!\approx\!\Delta t$),
\begin{equation}
\sigma_{E_I}(\Delta t)\ \gtrsim\ \frac{1}{\kappa}\sqrt{\frac{\overline{j}_A}{|A|\,\Delta t}} .
\label{eq:c6-var-EI-shot}
\end{equation}
Thus the \emph{force estimate} has a windowed cloud whose width decreases only as $\Delta t_{\!\mathrm{eff}}^{-1/2}$
unless $|A|$ or the collected inward flux $J_A$ is increased.

\paragraph{Finite correlation time.}
If arrivals have an effective correlation time $\tau>0$ (dead time, coherence, batching), the number
of independent samples in $\Delta t$ is $\approx \Delta t/\tau$, giving
\begin{equation}
\mathrm{Var}(\widehat{\overline{j}}_A)\ \gtrsim\ \frac{\overline{j}_A}{|A|}\,\frac{\tau}{\Delta t},
\qquad
\sigma_{E_I}\ \gtrsim\ \frac{1}{\kappa}\sqrt{\frac{\overline{j}_A}{|A|}\,\frac{\tau}{\Delta t}} .
\label{eq:c6-corr}
\end{equation}
This matches \eqref{eq:c6-crb-jbar}; see also \eqref{eq:c6-witness-jbar}. The identification
$\Delta t_{\!\mathrm{eff}}\!\sim\!\Delta t/\tau$ holds up to order‑one factors set by the measured autocorrelation.

\paragraph{Spherical screen (3D, far‑field).}
In the far field, for a full screen $A=4\pi r^2$ around a steady source with inward flux $J_{4\pi}$,
\begin{equation}
\overline{j}_A=\frac{J_{4\pi}}{4\pi r^2},
\qquad
\sigma_{E_I}\ \gtrsim\ \frac{1}{\kappa}\,\frac{1}{4\pi r^2}\,\sqrt{\frac{J_{4\pi}}{\Delta t}} .
\label{eq:c6-spherical}
\end{equation}
Directional diaphragms use the aperture area for $|A|$ and replace $J_{4\pi}$ by the collected inward flux.

\paragraph{Measured sectors (one‑datum calibration).}
With the sector mappings (one‑datum calibration; see \S\ref{subsec:op-calibration} and App.~\ref{app:calibrations})
\begin{equation}
\mathbf g=c_M\mathbf E_I,\qquad \mathbf E=c_Q\mathbf E_I,
\label{eq:c6-sector-maps}
\end{equation}
the estimator widths propagate as
\begin{equation}
\sigma_{g}\ \gtrsim\ \frac{c_M}{\kappa}\sqrt{\frac{F_A\,\overline{j}_A}{|A|\,\Delta t_{\!\mathrm{eff}}}},
\qquad
\sigma_{E}\ \gtrsim\ \frac{c_Q}{\kappa}\sqrt{\frac{F_A\,\overline{j}_A}{|A|\,\Delta t_{\!\mathrm{eff}}}} .
\label{eq:c6-measured}
\end{equation}
(Shot‑noise limit: replace $F_A\to1$ and $\Delta t_{\!\mathrm{eff}}\to\Delta t$.) An acceleration estimate
$a(r)=|\mathbf g|$ inherits the same $\Delta t_{\!\mathrm{eff}}^{-1/2}$ floor.

\paragraph{Trit‑density picture (screen shot‑noise).}
For a spherical screen with \emph{trit density} $N(r)=\alpha\,4\pi r^2$ and mean per‑trit hit rate $f$ (1/time),
the mean inward count in $\Delta t$ is $\mu=f\,N(r)\,\Delta t$ and
\[
\frac{\sigma_{\text{counts}}}{\mu}\ \approx\ \frac{1}{\sqrt{\mu}}
\ =\ \frac{1}{\sqrt{f\,\alpha\,4\pi r^2\,\Delta t}} .
\]
Mapping counts to $\mathbf E_I$ through \eqref{eq:c6-var-EI-shot} reproduces the same $1/\sqrt{|A|\Delta t}$ scaling.

\paragraph{Relation to time--information and capacity.}
The counting term in the C5 rate bound (\S\ref{corollary:time-info}) is precisely
$\overline{J}_{A}=\int_A(\mathbf j\!\cdot n)_-\,dA/\Delta t$. Equations \eqref{eq:c6-var-EI}–\eqref{eq:c6-corr}
give the corresponding estimator width for $\mathbf E_I$, showing that shorter windows both require more informational
flux (C5) and enlarge the force cloud. In applications that form durable records and later erase/compress them, the $P/T$
resource in C5 provides an additional operational limit; the variance floor here arises from counting statistics.

\paragraph{Noise kernel link (App.\ E).}
The local bounds above define a minimal \emph{noise kernel} for flux fluctuations on the boundary. App.~\ref{app:noise-kernel}
uses such kernels $\mathcal{N}$ to propagate covariances through linear response (Einstein–Langevin–type) and to connect
static shot noise to dynamical spectra.

\paragraph{Units and practical levers.}
All fluxes here are in nats/time (or \emph{trits}/time via $1/\ln 3$).
To tighten the cloud: enlarge $|A|$, increase the inward flux $J_A$ (e.g., by wider apertures or focusing),
extend $\Delta t$, reduce correlation time $\tau$, or multiplex independent patches and average.

\medskip
\noindent\emph{Summary.}
Windowed counting with or without finite correlation time implies an operational floor
\(\sigma_{E_I}\propto \sqrt{\overline{j}_A/(|A|\Delta t_{\!\mathrm{eff}})}\).
This sets a principled ``force cloud'' that narrows only with larger area, greater inward flux, or longer
integration time, and it propagates to calibrated sectors by a single one‑datum mapping.

\medskip
\noindent\emph{Literature note.}
Background on Poisson/renewal counting and shot noise: Kingman~\cite{Kingman1993}.
Stochastic/Langevin and spectral viewpoints: Gardiner~\cite{Gardiner2004}.
Appendix~\ref{app:noise-kernel} records the windowed noise‑kernel and covariance‑propagation formulas.
