% ======================================================================
% 04_operationalization.tex — Section 4: Operationalization & Calibration
% ======================================================================

\section{Operationalization \& Calibration}
\label{sec:operationalization}

This section records minimal steps to turn the Stokes identity on a region
$U\subset M$ with boundary $\partial U$ into numerical predictions for a chosen
measured sector. The spine is: generalized Stokes on $U$ and
on $U\times[t_0,t_1]$ (see \eqref{eq:stokes-identity}–\eqref{eq:cut-identity}),
the static linear closure \eqref{eq:static-closure}, and (when used) a
\emph{wave transport closure} recorded in \S\ref{sec:waves}
that provides a Poynting identity on source–free segments
(App.~\ref{app:forms-stokes}, eq.~\eqref{appB:eq:poynting}).

% --- Thermodynamic dictionary (boundary-first) ---
\paragraph{Thermodynamic dictionary (boundary-first).}
Piston face area $A$ $\to$ screen $S^{n-1}(r)$ with area $A_{n}(r)$;
pressure $p$ $\to$ normal flux density $j_n$ on the screen;
volume $V$ $\to$ enclosed $V_n(r)$;
mechanical work $\delta W$ $\to$ $j_n$ times the boundary deformation $A_{n}(r)\,\delta r$;
heat $\delta Q$ $\to$ boundary form $T\,\delta S_I$ (one datum per sector).
This dictionary is used operationally in C0 and C1 (\S\ref{corollary:info-gas}, \S\ref{corollary:static-screen}).

% --- Geometry note for n+1 and moving screens ---
\paragraph{Geometry in $n{+}1$ (screens and worldtubes).}
In $n$ spatial dimensions use the spherical surface $A_{n}(r)$ (or the effective cross‑section/circumference under confinement; C8: \S\ref{corollary:dimensional}).
For static tube/screen arguments work on an \emph{open annulus} $W\subset U$ with $\sigma|_W\equiv0$.
For moving screens replace $A_{n}(r)$ by the instantaneous $A_{n}(r,t)$ and use the worldtube \emph{cut (cap+side) identity} with side integrand $\mathbf j\!\cdot\!\mathbf n - i\,v_n$ (App.~\ref{app:forms-stokes}).

% --- Operational witnesses (local, optional) ---
\paragraph{Operational witnesses (local, optional).}
\emph{W-FluxConst[$\varepsilon$]}: on a static annulus with negligible production, the directed screen flux is invariant across nested screens (C1, \S\ref{corollary:static-screen}).
\emph{W-Narrow[$\varepsilon$]}: on a propagation segment, exported power is spectrally concentrated near a carrier (\S\ref{sec:waves}).
\emph{W-Lin[$\varepsilon$]}, \emph{W-Stat[$\varepsilon$]}, \emph{W-Ret[$\varepsilon$]}: local linearity, stationarity, and retardedness checks on the window (used when forming transfer functions).
Consequences in C1--C4 are invoked on windows where the corresponding witness is observed.

% ----------------------------- 4.1 -------------------------------------
\subsection{Inputs and outputs}
\label{subsec:op-inputs}

To make the framework predictive, the observer supplies:
\begin{enumerate}\itemsep2pt
  \item \textbf{Boundary choice.} A concrete $\partial U$: closed \emph{screen} or open \emph{sector/aperture}
        (cf.\ C0, \S\ref{corollary:info-gas}; C1, \S\ref{corollary:static-screen}). Confinement choices
        (guides/sheets) produce effective lower dimensions (C8, \S\ref{corollary:dimensional}).
  \item \textbf{Channel class.} Static (use $\mathbf j=-\kappa\,\mathbf E_I$) or a \emph{wave
        segment} (use the general wave template in \S\ref{sec:waves}; % <<< updated
        the Poynting identity \eqref{appB:eq:poynting} applies on source–free segments).
        If waves, state the constitutive description (e.g.\ $\chi$) on the segment where it is treated as homogeneous.
  \item \textbf{One datum (sector scale).} A single measurement $Y_\star$ on a reference
        screen/sector (radius $r_\star$, area $A_\star$) for the chosen readout
        $Y\in\{g,E,j_{\rm th}\}$; see \S\ref{subsec:op-calibration}.
  \item \textbf{Resource budgets.} Area $A$, time window $\Delta t$, power budget $P$,
        boundary temperature $T$ (\textbf{Kz}; App.~\ref{app:kick}), and a target reliability $\varepsilon$
        for coding/decoding.
  \item \textbf{Noise model (optional).} If needed, a noise kernel $\mathcal N$ for
        fluctuations (App.~\ref{app:noise-kernel}) or simply the Poisson shot–noise assumption
        used in C6 (\S\ref{corollary:noise}).
\end{enumerate}
\emph{Outputs} are calibrated fields and bounds over the same region and window:
static profiles $Y(r)$, wave envelopes $\langle|Y(r)|\rangle$, and operational
limits from time–information and noise.

% ----------------------------- 4.2 -------------------------------------
\subsection{One–datum calibration (constant–free scale setting)}
\label{subsec:op-calibration}

On a time–frozen, source–free annulus the directed flux $\Phi$ across any screen is
constant and
\[
\big\langle j_n\big\rangle(r)=\frac{\Phi}{A_n(r)},\qquad
|\mathbf E_I(r)|=\frac{1}{\kappa}\,\frac{\Phi}{A_n(r)}
\quad\text{(C0–C1; \S\ref{corollary:info-gas}, \S\ref{corollary:static-screen}).}
\]
A measured sector is a linear readout of $\mathbf E_I$:
\[
\mathbf g=c_M\,\mathbf E_I,\qquad
\mathbf E=c_Q\,\mathbf E_I,\qquad
\mathbf j_{\rm th}=\kappa_T\,\mathbf E_I.
\]
Fix $c_{\rm sec}$ by a single datum $Y_\star$ on a reference screen $(r_\star,A_\star)$
with the same flux $\Phi$:
\begin{equation}
c_{\rm sec}
=\frac{Y_\star}{|\mathbf E_I(r_\star)|}
=Y_\star\,\frac{\kappa\,A_\star}{\Phi}.
\label{eq:op-cal-const}
\end{equation}
Thereafter, in the lossless snapshot regime,
\begin{equation}
Y(r)=c_{\rm sec}\,|\mathbf E_I(r)|
=Y_\star\,\frac{A_\star}{A_n(r)}.
\label{eq:op-one-datum-ratio}
\end{equation}
If a flux label $\mathcal Q$ is used operationally (e.g.\ mass/charge within a source
family) with $\Phi\propto\mathcal Q$, then $Y(r)=Y_\star(\mathcal Q/\mathcal Q_\star)\,
[A_\star/A_n(r)]$ (App.~\ref{app:calibrations}).

\paragraph{Cones and confinement.}
For a conical sector of solid angle $\Delta\Omega$, replace $A_n(r)$ by
$A_\Delta(r)=\Delta\Omega\,r^{\,n-1}$. For guides/sheets use the effective constant
cross–section/circumference (C8, \S\ref{corollary:dimensional}). No new constants are introduced.

% ----------------------------- 4.3 -------------------------------------
\subsection{Wave segments: envelopes and export}
\label{subsec:op-waves}

On homogeneous source–free segments, the wave template in \S\ref{sec:waves} plus the
Poynting identity (App.~\ref{app:forms-stokes}, \eqref{appB:eq:poynting}) implies constant exported power
through screens. Time–averaging gives
\[
P(r)=\int_{\Sigma(r)}\!\langle \mathbf S\rangle\!\cdot n\,dA=\text{const},\qquad
\langle \mathbf S\rangle\propto \langle|\mathbf E_I|^2\rangle,
\]
hence the far–field envelope
\begin{equation}
\big\langle |\mathbf E_I(r)| \big\rangle \propto \frac{1}{\sqrt{A_n(r)}}
\quad\Rightarrow\quad
\big\langle |Y(r)| \big\rangle = c_{\rm sec}\,\big\langle |\mathbf E_I(r)| \big\rangle .
\label{eq:op-wave-envelope}
\end{equation}
In short time windows $\Delta t$, interference can suppress instantaneous export on sub‑windows,
while the \emph{window‐integrated} export $\int_t^{t+\Delta t}\!P(\tau)\,d\tau$ remains radius‑independent
on the same segment. Resolution is bounded by \eqref{eq:op-rate}–\eqref{eq:op-time} and the noise floor
\eqref{eq:op-noise}. Confinement replaces $A_n(r)$ by the effective cross–section/circumference (C8,
\S\ref{corollary:dimensional}). Apply C11 corrections if attenuation or capacity ceilings are relevant
(C11, \S\ref{corollary:entropic-corrections}).

% ----------------------------- 4.4 -------------------------------------
\subsection{Operational limits: time, capacity, and erasure}
\label{subsec:op-limits}

Let $\overline{J}_{A}$ be the inward counting rate over a patch $A$ in a window
$\Delta t$ (C5, \S\ref{corollary:time-info}). With a reliability target $\varepsilon$
and a channel class choice (coding, alphabet, constraints), let
$C_{\rm chan}^{(\mathrm{trits})}(\varepsilon)$ denote the corresponding $\varepsilon$–reliable
capacity (trits per time). Then the resource–limited bounds (C5) are
\begin{equation}
\frac{\Delta I_{\rm trits}}{\Delta t}
\le \min\!\left\{\frac{\overline{J}_{A}}{\ln 3},\
C_{\rm chan}^{(\mathrm{trits})}(\varepsilon),\
\frac{P}{T\,\ln 3}\right\},
\label{eq:op-rate}
\end{equation}

\begin{equation}
\Delta t
\ge \max\!\left\{\frac{\Delta I_{\rm trits}}{\overline{J}_{A}/\ln 3},\
\frac{\Delta I_{\rm trits}}{C_{\rm chan}^{(\mathrm{trits})}(\varepsilon)},\
\frac{\Delta I_{\rm trits}\,\ln 3\,T}{P}\right\}.
\label{eq:op-time}
\end{equation}

\paragraph{Erasures/unknowns (resolved content).}
If a fraction $\varepsilon_{\bot}\in[0,1)$ of symbols observed on the same window are erasures $(\bot)$,
and one wishes to bound \emph{resolved} content only, make the substitution
\begin{equation}
\overline{J}_{A}\ \mapsto\ (1-\varepsilon_{\bot})\,\overline{J}_{A},
\qquad
C_{\rm chan}^{(\mathrm{trits})}\ \mapsto\ (1-\varepsilon_{\bot})\,C_{\rm chan}^{(\mathrm{trits})}
\end{equation}
in \eqref{eq:op-rate}–\eqref{eq:op-time}. The $P/T$ term is unchanged.

% ----------------------------- 4.5 -------------------------------------
\subsection{Noise floor and estimator width (operational counting witness)}
\label{subsec:op-noise}

Let $N_A(\Delta t)$ be the inward counts on a patch $A$ over a window $\Delta t$.
Define the \emph{measured} mean inward current
\[
\overline{j}_A:=\frac{\mathbb E[N_A]}{|A|\,\Delta t},
\]
the windowed \emph{Fano factor} \(F_A:=\mathrm{Var}(N_A)/\mathbb E[N_A]\), and an
\emph{effective window} \(\Delta t_{\!\mathrm{eff}}\le \Delta t\) obtained from the
count autocorrelation on the same record (App.~\ref{app:noise-kernel}).
Any boundary estimator \(Y\) built from these counts admits the \emph{operational}
benchmark width
\begin{equation}
\sigma_{Y}\ \approx\ \frac{c_{\rm sec}}{\kappa}\,
\sqrt{\frac{F_A\,\overline{j}_A}{|A|\,\Delta t_{\!\mathrm{eff}}}}.
\label{eq:op-noise}
\end{equation}
Shot–noise–limited records have \(F_A\!\approx\!1\) and \(\Delta t_{\!\mathrm{eff}}\!\approx\!\Delta t\);
overdispersed or correlated arrivals are captured by the measured \((F_A,\Delta t_{\!\mathrm{eff}})\).
This is a windowed, data–driven estimate.

% ----------------------------- 4.6 -------------------------------------
\subsection{Procedure checklist (static and wave cases)}
\label{subsec:op-checklist}

\noindent\emph{Static.}
\begin{enumerate}\itemsep2pt
  \item Choose $\partial U$ (screen/sector) and, if needed, confinement geometry (C8, \S\ref{corollary:dimensional}).
  \item Apply \eqref{eq:stokes-identity} and the \emph{cut identity} \eqref{eq:cut-identity} with $\sigma|_W\equiv0$ on an open annulus $W\subset U$ to identify
        a constant directed flux $\Phi$.
  \item Fix $c_{\rm sec}$ from one datum $Y_\star$ on $(r_\star,A_\star)$ via \eqref{eq:op-cal-const}.
  \item Predict $Y(r)$ from the ratio law \eqref{eq:op-one-datum-ratio}; apply C11 corrections if attenuation
        or capacity ceilings are relevant (C11, \S\ref{corollary:entropic-corrections}).
  \item If a finite window/area is used, report estimator width from \eqref{eq:op-noise}.
\end{enumerate}

\noindent\emph{Waves.}
\begin{enumerate}\itemsep2pt
  \item Specify the segment and constitutive description where it is treated homogeneous; ensure sources are excluded (\S\ref{sec:waves}).
  \item Use the Poynting identity to obtain constant power $P(r)$ and the envelope
        \eqref{eq:op-wave-envelope}; apply confinement if appropriate (C8, \S\ref{corollary:dimensional}).
  \item Calibrate with the same $c_{\rm sec}$ and quantify bounds via
        \eqref{eq:op-rate}–\eqref{eq:op-time} for the chosen reliability $\varepsilon$.
  \item Apply C11 corrections if attenuation or capacity ceilings are relevant (C11, \S\ref{corollary:entropic-corrections}).
\end{enumerate}

% ----------------------------- 4.7 -------------------------------------
\subsection{Micro–example (3D static screen)}
\label{subsec:op-example}

A spherical screen of radius $r$ in $n=3$ has area $4\pi r^2$.
If $\Phi$ is the measured tube flux on a static annulus with negligible production,
\begin{equation}
\langle j_n(r)\rangle=\frac{\Phi}{4\pi r^2} .
\end{equation}
With the snapshot closure \eqref{eq:static-closure}, the informational field amplitude scales as
\begin{equation}
\big|\mathbf E_I(r)\big|=\frac{1}{\kappa}\,\frac{\Phi}{4\pi r^2}.
\end{equation}
This furnishes the static $1/r^2$ profile used in examples.

% ----------------------------- 4.8 -------------------------------------
\subsection{Units (Kz) and trits/nats}
\label{subsec:op-units}

Kz units are used for rates; base logs convert by $\ln 3$:
\begin{equation}
S_{\rm nats}=(\ln 3)\,S_{\rm trits}.
\end{equation}
Formulas above are invariant under the trit/nat choice except for the explicit factors
of $\ln 3$ in the time–information bounds \eqref{eq:op-rate}–\eqref{eq:op-time}.
(Recovering binary units replaces $\ln 3$ by $\ln 2$ only.)

\medskip
\noindent\emph{Literature note.}
Capacity and reliability follow Shannon and Gallager~\cite{Shannon1948,Gallager1968};
erasure cost follows Landauer and Bennett~\cite{Landauer1961,Bennett2003};
for wave transport/export see \S\ref{sec:waves} and App.~\ref{app:forms-stokes};
for boundary counting statistics, the windowed Fano factor and autocorrelation–based $\Delta t_{\!\mathrm{eff}}$ are standard empirical tools
(App.~\ref{app:noise-kernel}).
